Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis

### 2022-11-26 analysis/paper/paper.Rmd
edit code chunk

---
title: "Investigating Neolithic pottery use in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the use of pottery vessels at Guijiabao. Lipid analysis have proven useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. Lipid residues are the most direct evidence to investigate culinary practices and dietary patterns since specific lipid compounds can be associated with different types of foodstuffs cooked or stored in pottery vessels [@Copley2005]. It is especially informative for human subsistence in the situation where archaeological sites lack organic remains due to acidic sediments [@Kwak2020]. The growing application of organic residue analysis demonstrate it offers a better understanding of culinary practices during the transitional period to agricultural societies [@Craig2011], dietary patterns in

@@

## Samples
We selected five ceramic fragments (Figure XX) from Guijiabao for organic residue analysis, in which four samples were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season. To further investigate the pottery use in an earlier period, we focus on pottery vessels from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.). During this period, the common vessel forms include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are mostly fragments from jars of those two major types that were supposed to be used for food serving or storage according to the archaeological context of living area [].

@@

# Results
Lipid residues from our pottery samples show that the saturated medium-chain fatty acids, palmitic (C16:0) and stearic(C18:0), are the two dominant compounds. Those two compounds were normally degraded from animal fat and plant wax that indicates the organic compounds in our samples were highly degraded

. However, for the sample five, we detected rich compounds with high intensity compared to the other four samples. This might indicate a different source.

To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples. Here you can see the stable carbon isotope measurements of C16:0 and C18:0 fatty acids. The results suggest a dominance of C3 plant for most pots, and a few shows a mixture of animal fats and C3 plants. The sample 5 where more compounds were preserved shows a source of C4 plant or marine animals instead. We think this is very interesting finding since there was just a few rice grains found at this site compared to the evidence of millets.

>>>>

---
title: "Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the vessel usage at Guijiabao. Lipid analysis have proven useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. Lipid residues are the most direct evidence to investigate culinary practices and dietary patterns since specific lipid compounds can be associated with different types of foodstuffs cooked or stored in pottery vessels [@Copley2005]. It is especially informative for human subsistence in the situation where archaeological sites lack organic remains due to acidic sediments [@Kwak2020]. The growing application of organic residue analysis demonstrate it offers a better understanding of culinary practices during the transitional period to agricultural societies [@Craig2011], dietary patterns in

@@

## Samples
We selected five ceramic fragments (Table XX) from Guijiabao for organic residue analysis, in which four samples were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season. To further investigate the content and use of pottery vessels in an earlier period, we focus on potsherds excavated from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.). During this period, the common vessel forms include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are jar fragments from those two common types supposed to be used for food serving or storage according to their archaeological context that suggests living area [].

@@

# Results

```{r tidy data}
library(tidyverse)

# load the GC data
meth <- read_csv(here::here("analysis/data/raw_data/LY_21_08_20_pot_chrom.CSV"))

# data wrangling
ixd <- which(str_detect(meth$Path, "C"))
meths <- split(meth, cumsum(1:nrow(meth) %in% ixd))
meth_data <- meths[c(FALSE, TRUE)]
meth_name <- meths[c(TRUE, FALSE)]
names(meth_data) <- map(meth_data, ~.x$Path[1])
names(meth_name) <-map(meth_name, ~.x$Sample[1])
meths <- map_df(meth_data, ~.x[-1, ] %>%
                   mutate_all(as.numeric), .id = 'sample')

# add sample name
meth_all <-
  meths %>%
  mutate(sample = paste0(parse_number(sample),
                         str_extract(sample, pattern = "_00[0-9]"))) %>%
  mutate(name = case_when(
    sample %in% c("210729_001") ~ "CDG-062",
    sample %in% c("210729_002") ~ "SYG-3",
    sample %in% c("210729_003") ~ "SYG-TN13-E22-2#1",
    sample %in% c("210729_004") ~ "SYG-TN13-E22-1",
    sample %in% c("210729_005") ~ "SYG-TN13-E22-2#2",
    sample %in% c("210729_006") ~ "SYG-TN13-E23-3")) %>%
  mutate(name = factor(name, levels = c("SYG-TN13-E22-1",
                                        "SYG-TN13-E22-2#1",
                                        "SYG-TN13-E22-2#2",
                                        "SYG-TN13-E23-3",
                                        "SYG-3",
                                        "CDG-062")))
```

Lipid residues from our samples show that the saturated fatty acids with even number of carbons (C14:0, C16:0, and C18:0), particularly palmitic (C16:0) and stearic(C18:0), are the dominant compounds. Because palmitic and stearic fatty acids are normally degraded from animal fat or plant wax, the domination pattern with little concentration of other fatty acids suggests the organic compounds in our samples were highly degraded during post depositional processes. However, there is one sample contains rich compounds that include long-chain fatty acids, C20, C22, and C24, which were absent in other samples. This might indicate

To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples. Here you can see the stable carbon isotope measurements of C16:0 and C18:0 fatty acids. The results suggest a dominance of C3 plant for most pots, and a few shows a mixture of animal fats and C3 plants. The sample 5 where more compounds were preserved shows a source of C4 plant or marine animals instead. We think this is very interesting finding since there was just a few rice grains found at this site compared to the evidence of millet.
